---
title: "GEDI Level 1B Waveforms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GEDI Level 1B Waveforms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options("chewie.testing" = TRUE)
temp_dir <- tempdir()

if (!file.exists(".Renviron")) {
  file.create(".Renviron")
}

cache_zip <- system.file(
  "chewie-test-cache.zip",
  package = "chewie"
)

utils::unzip(
  cache_zip,
  exdir = temp_dir
)
temp_cache <- file.path(temp_dir, "chewie-test-cache")

chewie::chewie_setup_cache(temp_cache, renviron = "project")
```

This vignette demonstrates how to use the `chewie` package to download and
extract GEDI Level 1B waveforms. 

First let's load the `chewie` package and the `ggplot2` package for plotting.

```{r setup}
suppressPackageStartupMessages({
  library(chewie)
  library(ggplot2)
  library(sf)
  library(dplyr)
})
```

First, we define an area of interest. In this case, we will use prairie creek
national park as our area of interest. The `chewie` package provides a number
of example spatial vector data sets that can be used for this purpose. Here we
use the `prairie-creek.geojson` data set, which is a simple polygon defining
the park boundary. You can alternatively use the `st_read()` function to
read in your own spatial vector data.

```{r define-aoi}
pcreek <- system.file(
  "geojson", "prairie-creek.geojson",
  package = "chewie"
) |>
  sf::read_sf()
```

Now we can search for GEDI Level 1B products that intersect with our area of
interest. We will search for products that were collected between January 1st
and January 31st, 2023.

```{r find-gedi}
g1b_search <- find_gedi(pcreek,
  gedi_product = "1B",
  date_start = "2023-01-01",
  date_end = "2023-01-31"
)
print(g1b_search)
```
You'll see that only one granule was found. The returned `gedi.find` object 
contains information about the search, including whether or not the files are 
cached locally.

Now we can download the GEDI Level 1B products. However, note that the `grab_gedi()`
function will not download the products if they are already cached locally. It 
first checks the cache and grabs the relevant files if they exist. If they do
not exist, they'll be downloaded from the LPCLOUD bucket. {chewie} will, by 
default only download specific data columns. There are many variables contained
in the raw hdf5 files.

```{r grab-gedi}
gedi_1b_arrow <- grab_gedi(g1b_search)
```

The  `grab_gedi()` function returns a *arrow_dplyr_query* object. If the user 
wishes to carry out further filtering (for example using quality flags), this 
should be done at this stage to avoid loading unnecessary data into memory.

In order to access the data we must first collect the data into a data frame.
This could be done using `dplyr::collect`. However, the `collect_gedi()` function
is a wrapper around `dplyr::collect` that also converts the data to an `sf`
object, immediately providing a spatial context for the data. 
```{r plot-gedi}
g1b_sf <- collect_gedi(gedi_1b_arrow, g1b_search)

print(g1b_sf)

plot(g1b_sf[0], axes = TRUE)
plot(attributes(g1b_search)$aoi, add = TRUE, border = "#7700ff")
```

So, here's the cool bit about the way the Level-1B data are extracted. We can
access the full waveform for every shot in the data set by using the 
`extract_waveforms()` function. This generates a long-form data frame with 
waveform amplitude, relative elevation, and shot number for each shot. This enables
the combined analysis and comparison of many waveforms across space (and time 
when we have multiple granules). 

For example, here we take 8 of the waveforms...
```{r extract-waveforms}
g1b_wvf <- extract_waveforms(g1b_sf[100:108, ])

print(g1b_wvf)
```

And plot them using `ggplot2`...

```{r plot-waveforms}
wvf_ggplot <- function(x, wf, z, .ylab = "Elevation (m)") {
  wf <- sym(wf)
  z <- sym(z)
  ggplot(x, aes(x = !!z, y = !!wf)) +
    geom_ribbon(aes(ymin = min(!!wf), ymax = !!wf),
      alpha = 0.6, fill = "#69d66975", colour = "grey30", lwd = 0.2
    ) +
    theme_light() +
    coord_flip() +
    labs(x = .ylab, y = "Waveform Amplitude")
}

wvf_ggplot(g1b_wvf, "rxwaveform", "rxelevation") +
  facet_wrap(~shot_number, ncol = 5, scales = "free")
```

We can also calculate the mean waveform for these shots and plot it. *Note that
this may not be particualtly helpful for all contexts, but it provides an indication
of the power of having access to all waveforms in a single dataset.*

```{r plot-mean-waveforms}
g1b_wvf |>
  group_by(shot_number) |>
  arrange(rxelevation) |>
  mutate(id = dplyr::n():1) |>
  ungroup() |>
  group_by(id) |>
  summarise(
    mean_wvf = mean(rxwaveform),
    mean_elev = mean(rxelevation)
  ) |>
  wvf_ggplot("mean_wvf", "id", .ylab = "Sequence order") +
  scale_x_reverse()
```

```{r cleanup}
if (file.exists(".Renviron")) {
  file.remove(".Renviron")
}
```